```{r}
library(tidyverse)
set.seed(222)
theme_set(theme_bw())
```

Here we'll verify our understanding of a statistical procedure using simulation.

## 1. Choose the parameters for your simulation

```{r}
p1 <- 0.60
n1 <- 25
p2 <- 0.52
n2 <- 25
```

## 2. Simulate a sample and calculate the sample statistic

Use `rbinom()` for binary response variables (like Soil Pb > 200).

```{r}
pre1950 <- rbinom(n = 1, size = n1, prob = p1)
post1950 <- rbinom(n = 1, size = n2, prob = p2)

c(pre1950, post1950)

pre1950 / n1 - post1950 / n2
```

## 3. Continue simulating samples until you reach critical mass

```{r}
pre1950 <- rbinom(n = 1e4, size = n1, prob = p1)
post1950 <- rbinom(n = 1e4, size = n2, prob = p2)

phat1 <- pre1950 / n1
phat2 <- post1950 / n2
diff_props <- phat1 - phat2

head(diff_props)
tibble(diff_props) |>
  ggplot(aes(diff_props)) +
  geom_histogram(binwidth = 0.04) +
  geom_vline(xintercept = p1 - p2, color = "cornflowerblue", linewidth = 1.2)
```

## 4. Check your simulation results against the claim 

_Do 95% of sample statistics fall within 1.96 SEs of the population parameter?_

```{r}
se <- sqrt(phat1 * (1 - phat1) / n1 + phat2 * (1 - phat2) / n2)
head(se)
mean(p1 - p2 >= diff_props - 1.96 * se & p1 - p2 <= diff_props + 1.96 * se)
```

Pretty good! But not perfect. Notice the coverage is consistently ~93.8%, which is slightly off from 95%. There are a couple contributing factors to this small, but consistent, difference. 

I'll leave you with two closing questions for you to ponder.

1. What do you think might be affecting the accuracy of the inference?
2. Once you come up with a hypothesis or two, how would you modify this experiment to test your hypothesis?
