```{r}
library(tidyverse)
set.seed(222)
theme_set(theme_bw())
```

We're going to see how to sample from both *empirical* and *parametric* distributions in R. This is the foundation of *simulation*, which is the most important tool in this course.

## Empirical distributions

Read empirical data and visualize the whole distribution.

```{r}
soil_cores <- read_csv("data/exide-soil-cores.csv")

# Visualize the whole distribution
ggplot(soil_cores, aes(pb_ppm)) +
  geom_histogram(binwidth = 20) +
  xlim(0, 1000)
```

`sample()` draws samples from a vector (e.g. a column).

```{r}
pb_ppm_sample <- sample(soil_cores$pb_ppm, size = 50)

tibble(pb_ppm_sample) |>
  ggplot(aes(pb_ppm_sample)) +
  geom_histogram(binwidth = 50)
```

`slice_sample()` draws samples from a data frame.

```{r}
soil_sample <- soil_cores |>
  slice_sample(n = 50)

ggplot(soil_sample, aes(pb_ppm)) +
  geom_histogram(binwidth = 50)
```

## Parametric distributions

Visualize the PDF with `dnorm()` and other `d____()` functions. 

```{r}
tibble(
  x = seq(-860, 1366, length.out = 1000),
  pdf = dnorm(x, mean = 253, sd = 371)
) |>
  ggplot(aes(x, pdf)) +
  geom_line(linewidth = 1.5)
```

Draw samples with `rnorm()` and other `r____()` functions.

```{r}
normal_sample <- tibble(x = rnorm(25, mean = 253, sd = 371))
ggplot(normal_sample, aes(x)) +
  geom_histogram(binwidth = 100)
```
