```{r}
library(tidyverse)
set.seed(222)
theme_set(theme_bw())
```

Simulate data following a linear regression model

Our model in statistical notation is:

$$
\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon \\
\epsilon &\sim Normal(0, \sigma)
\end{align}
$$

For this example, we'll _simulate_ the relationship between water temperature and the length of copepods.

1. Choose parameters
2. Simulate data
3. Fit the model
4. Interpret the model

## Choose parameters

Our three model parameters are the two $\beta$s and $\sigma. We also need to define our sample size, $n$. 

```{r}
beta_0 <- 3000 # intercept: length of a copepod (in um) when temp is 0 degC
beta_1 <- -55 # slope: change in length per degC
sigma <- 100 # standard deviation of the residuals, the "spread" around the line
n <- 50 # sample size
```

## Simulate data

1. Simulate the _predictor_
2. Calculate the _expected_
3. Simulate the _response_
4. Assemble the data frame

### Simulate the predictor

For temperature, _uniformly_ distributed random numbers between 2 and 14 degC.

```{r}
temp_c <- runif(n, min = 2, max = 14)
```

### Calculate the expected

The value of the line. $y$ without residuals, $\epsilon$.

$$
y = \beta_0 + \beta_1 x
$$


```{r}
expected_length_um <- beta_0 + beta_1 * temp_c
```

### Simulate the response

The response is _normally distributed_ around the line, thanks to the residuals.

```{r}
length_um <- rnorm(n, mean = expected_length_um, sd = sigma) # notice the mean is the _vector_ of expected values
```

### Assemble the data frame

```{r}
# Create a tibble
copepods <- tibble(temp_c, length_um)

# Visualize the data
ggplot(copepods, aes(x = temp_c, y = length_um)) +
  geom_point()
```

## Fit the model

Use `lm()` to fit a model. Specify the response and predictor with a _formula_.

```{r}
copepod_lm <- lm(length_um ~ temp_c, data = copepods)
```

Done and done!

## Interpret the model

Inspect the coefficients with `summary()`.

```{r}
summary(copepod_lm)
```

Note the estimates and their standard errors (SEs). 95% of the time, the estimates will be within 2 SEs of the population parameters (i.e., our chosen simulation values.)

Use `sigma()` to check the model estimate of sigma.

```{r}
sigma(copepod_lm)
```

Pretty close!

Our residuals should be normally distributed, with a sd of sigma.

```{r}
copepods |>
  # Get the residuals from the model
  mutate(residual = residuals(copepod_lm)) |>
  ggplot(aes(x = residual)) +
  geom_histogram()
```

