---
title: "Lab 3: Simulate data and fit a regression model"
subtitle: "Ordinary least squares"
format:
  html:
    other-links:
      - text: Download Source
        href: lab3.qmd
execute:
  eval: true
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
set.seed(222)
theme_set(theme_bw())
```

## Before you code

Before you start coding in this document, your group should have completed the sections **Design the scenario** and **Choose your parameters** from the handout.

## Simulate the data

You've chosen your parameters ($\beta_0$, $\beta_1$, and $\sigma$). Define them here:

```{r}
# These values chosen to represent the negative relationship between body size and temperature in marine invertebrates (copepods)
beta_0 <- 3000
beta_1 <- -55
sigma <- 100
```

### Generate random values

1. Simulate your predictor variable.

::: {.callout-tip}

You may use any random number generating function to generate your predictor variable. Often, it's simplest to simulate the predictor variable using the _uniform_ distribution. Here's an example of generating 5 random numbers between 1 and 2:

```{r}
runif(n = 5, min = 1, max = 2)
```

:::

```{r}
# In the range 2-14degC
temp_c <- runif(n = 30, min = 2, max = 14)
```

2. Calculate the expected value of the response.

```{r}
# Expected length in micrometers (um)
expected_length_um <- beta_0 + beta_1 * temp_c
```

3. Simulate the response variable.

```{r}
# The simulated lengths
length_um <- rnorm(
  n = 30,
  mean = expected_length_um,
  sd = sigma
)
```

Visualize your simulated data. Verify it's comparable to your original drawing. If necessary, tweak your parameter values until it represents your scenario reasonably well.

```{r}
# First, put your predictor and response variables into a single data frame.
simulated_lengths <- tibble(temp_c, length_um)

# Then, use ggplot to make a scatter plot.
ggplot(simulated_lengths, aes(x = temp_c, y = length_um)) +
  geom_point()
```

## Fit and interpret a regression model

### Fit the model

Use `lm()` to fit a regression model to your simulated data. 


::: {.callout-tip}

`lm()` expects a _formula_ stating your predictor and response variables. It should look like: `response ~ predictor`.

:::


```{r}
# Fit and summarize model
length_model <- lm(length_um ~ temp_c, data = simulated_lengths)
summary(length_model)
sigma(length_model)
```

### Interpret the model

Refer back to the worksheet to complete most of this section. When you reach the end ("Test your intuition by modifying your simulation to increase $\sigma$ and n independently, then re-run it several times."), come back to this document and use the following code chunks to test your intuition.

**Increase $\sigma$**

```{r}
# Keep your predictor variable the same, but increase sigma by 2x
double_sigma <- 2 * sigma

# Modify the following lines of code to re-run your simulation with the bigger sigma
expected_length_um2 <- beta_0 + beta_1 * temp_c
length_um2 <- rnorm(
  n = 30,
  mean = expected_length_um,
  sd = double_sigma
)
simulated_lengths2 <- tibble(temp_c, length_um = length_um2)
length_model2 <- lm(length_um ~ temp_c, data = simulated_lengths2)

# Get your betas and their standard errors
summary(length_model2)

# How do they compare to the population parameters, relative to the original simulation?
```

My $\beta$ estimates only shifted a little bit (consistent with randomness) but the standard errors got much larger. Bigger residuals mean less certainty about the coefficient estimates.

**Increase n**

```{r}
# Now increase n by 2x. You'll need to create more predictor variable values.
temp_c_2n <- c(
  temp_c,
  runif(n = 30)
)

# Modify the following lines of code to re-run your simulation with the bigger n
expected_length_um3 <- beta_0 + beta_1 * temp_c_2n
length_um3 <- rnorm(
  n = 60,
  mean = expected_length_um3,
  sd = sigma
)
simulated_lengths3 <- tibble(temp_c = temp_c_2n, length_um = length_um3)
length_model3 <- lm(length_um ~ temp_c, data = simulated_lengths3)

# Get your betas and their standard errors
summary(length_model3)

# How do they compare to the population parameters, relative to the original simulation?
```

Once again, $\beta$ estimates remained pretty similar, but this time the standard error _decreased_. Bigger sample size means less uncertainty!
