---
title: "EDS 222: Statistics for Environmental Data Science"
subtitle: "Week 3 Monday: Intro to OLS"
format: revealjs
execute:
    echo: false
---

<style>
.reveal pre.small-output code {
  font-size: 80%;
}
</style>

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(geosphere)
set.seed(222)
theme_set(theme_bw(22))

# Approximate location of the former Exide facility (2700 S Indiana St, Vernon)
exide_lon <- -118.1923
exide_lat <- 34.0058

soil_cores <- read_csv("data/exide-soil-cores.csv") |>
  filter(location_code == "YD", pb_ppm <= 5e3) |>
  mutate(
    distance_ft = distGeo(cbind(longitude, latitude), c(exide_lon, exide_lat)) *
      3.28,
    distance_mi = distance_ft / 5280,
    bearing_deg = bearing(c(exide_lon, exide_lat), cbind(longitude, latitude))
  ) |>
  drop_na()

step_out <- read_csv("data/step-out.csv") |>
  mutate(distance_mi = distance_ft / 5280)

all_samples_north <- bind_rows(
  `soil core` = transmute(
    filter(soil_cores, between(bearing_deg, 0, 30)),
    property_id,
    distance_ft,
    distance_mi,
    pb_ppm
  ),
  `step out` = select(step_out, property_id, distance_ft, distance_mi, pb_ppm),
  .id = "source"
) |>
  drop_na()
```

# Intro to OLS

## Learning objectives

- **Describe relationships** between variables using a _regression model_
- **Estimate** _coefficients_ using _ordinary least squares_
- **Interpret** the _parameters_ of a regression model

# Describe relationships between variables using a regression model

## How much soil lead would you _expect_ to find 2 mi from Exide?

```{r}
#| fig-width: 6
#| fig-align: center
ggplot(all_samples_north, aes(x = distance_mi, y = pb_ppm)) +
  geom_point(shape = 21, alpha = 0.25) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  labs(x = "Distance (mi)", y = "Soil Pb (ppm)")
```

## Fitting a line to data

The line is our **model** explaining the relationship between two variables

$$
y = mx + b
$$

* $y$ is the _response_ (AKA dependent) variable
* $x$ is the _predictor_ (AKA independent) variable
* $m$ is the slope
* $b$ is the intercept

## In a world without uncertainty...

Without uncertainty, all the points would fall on the line. 


::: {.columns}

::: {.column}

```{r}
#| fig-width: 6
#| fig-align: center
lead_lm <- lm(pb_ppm ~ distance_mi, all_samples_north)
all_samples_north |>
  mutate(pb_ppm = fitted(lead_lm)) |>
  ggplot(aes(x = distance_mi, y = pb_ppm)) +
  geom_point(shape = 21, alpha = 0.25) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  ylim(0, 2200) +
  labs(x = "Distance (mi)", y = "Soil Pb (ppm)")
```

:::

::: {.column}

```{r}
#| fig-width: 6
#| fig-align: center
ggplot(all_samples_north, aes(x = distance_ft / 5280, y = pb_ppm)) +
  geom_point(shape = 21, alpha = 0.25) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  labs(x = "Distance (mi)", y = "Soil Pb (ppm)")
```

:::

:::

## Fitting a line to data

$$
\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon \\
\epsilon &\sim Normal(0, \sigma)
\end{align}
$$

In real world data, the data spread out around the line.

- The difference between the point and the line is the **residual**.
- Linear regression assumes the residuals are **normally distributed**.

## Fit a lead soil model

Let's fit a linear model to the lead soil data

```{r}
#| echo: true

# all_samples_north = all samples north of Exide
soil_lm <- lm(pb_ppm ~ distance_mi, all_samples_north)

coef(soil_lm)
```

$$
\text{Pb (ppm)} = 325.6 - 46.1 \times \text{Distance (mi)} + \epsilon
$$

## A closer look at ENA-31

ENA-31 is 1.13 mi from Exide. This soil sample had Pb 931 ppm.

```{r}
#| fig-width: 6
#| fig-align: center
ena_31 <- filter(all_samples_north, property_id == "ENA-31", pb_ppm > 500)
ggplot(all_samples_north, aes(x = distance_mi, y = pb_ppm)) +
  geom_point(shape = 21, alpha = 0.25) +
  geom_point(
    data = ena_31,
    color = "firebrick",
    size = 6,
    shape = 21,
    stroke = 3
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  labs(x = "Distance (mi)", y = "Soil Pb (ppm)")
```

## The _expected_ soil lead level

ENA-31 is 1.13 mi from Exide. This soil sample had Pb 931 ppm.

The **expected** value of soil Pb at that distance is:

$$
\begin{align}
\text{Pb (ppm)} &= 325.6 - 46.1 \times \text{Distance (mi)} \\
\text{Pb (ppm)} &= 325.6 - 46.1 \times 1.13 \\
\text{Pb (ppm)} &= 273.5 
\end{align}
$$

The **residual** is the difference between the observed (931) and the expected (273.5).

## _Residual_ soil lead

```{r}
#| fig-width: 6
#| fig-align: center
ggplot(all_samples_north, aes(x = distance_mi, y = pb_ppm)) +
  geom_point(shape = 21, alpha = 0.25) +
  geom_segment(
    aes(yend = 273.5, xend = distance_mi),
    ena_31,
    color = "firebrick",
    linewidth = 1.2
  ) +
  geom_point(
    data = ena_31,
    color = "firebrick",
    size = 6,
    shape = 21,
    stroke = 3
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  labs(x = "Distance (mi)", y = "Soil Pb (ppm)")
```

# Estimate coefficients using ordinary least squares

## Picking the right line

There are lots of possible lines through these data, how did the model find its **coefficients**? 

$\text{Pb (ppm)} = 325.6 - 46.1 \times \text{Distance (mi)}$

```{r}
#| fig-width: 6
#| fig-align: center
ggplot(all_samples_north, aes(x = distance_mi, y = pb_ppm)) +
  geom_point(shape = 21, alpha = 0.25) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  labs(x = "Distance (mi)", y = "Soil Pb (ppm)")
```

## OLS minimizes the _squared_ residuals

```{r}
#| fig-width: 6
#| fig-align: center
# Stratified random sample: 1 point from each of 8 equal-width distance bins
ols_sample <- all_samples_north |>
  mutate(stratum = cut(distance_mi, 8)) |>
  slice_sample(n = 1, by = stratum)
sample_lm <- lm(pb_ppm ~ distance_mi, ols_sample)

# coord_fixed() makes 1 ppm on the y-axis the same length as `ratio` mi on
# the x-axis, so a box `ratio * |residual|` wide is a square on screen
ratio <- diff(range(ols_sample$distance_mi)) /
  diff(range(ols_sample$pb_ppm)) *
  0.75
ols_sample <- ols_sample |>
  mutate(
    fitted = fitted(sample_lm),
    resid = residuals(sample_lm),
    sq_xmin = distance_mi,
    sq_xmax = distance_mi + abs(resid) * ratio,
    sq_ymin = pmin(pb_ppm, fitted),
    sq_ymax = pmax(pb_ppm, fitted)
  )

ggplot(ols_sample, aes(x = distance_mi, y = pb_ppm)) +
  geom_rect(
    aes(xmin = sq_xmin, xmax = sq_xmax, ymin = sq_ymin, ymax = sq_ymax),
    fill = "firebrick",
    alpha = 0.2
  ) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  geom_segment(
    aes(xend = distance_mi, yend = fitted),
    color = "firebrick",
    linewidth = 1.2
  ) +
  geom_point(shape = 21, size = 3, fill = "white") +
  coord_fixed(ratio = ratio) +
  labs(
    title = paste("SSR =", scales::comma(sum(ols_sample$resid^2))),
    x = "Distance (mi)",
    y = "Soil Pb (ppm)"
  )
```

## Different coefficients yield higher SSR 

```{r}
#| fig-width: 6
#| fig-align: center
# Reuses ols_sample, sample_lm, and ratio from the previous slide.
# Wander through candidate (beta_0, beta_1) pairs, ending at the OLS estimate
b_hat <- coef(sample_lm)
x_bar <- mean(ols_sample$distance_mi)
y_bar <- mean(ols_sample$pb_ppm)
# Each keyframe shifts the line's height at x_bar and tilts its slope
keyframes <- tribble(
  ~shift , ~tilt ,
     250 ,     0 ,
    -150 ,   120 ,
     150 ,   -80 ,
     -60 ,    30 ,
       0 ,     0
) |>
  mutate(
    b1 = b_hat[2] + tilt,
    b0 = y_bar + shift - b1 * x_bar
  )
ease <- function(t) (1 - cos(pi * t)) / 2
frames_per_hold <- 10
frames_per_move <- 15
# Pause on each keyframe so the SSR is legible, then glide to the next one
candidate_lines <- map(seq_len(nrow(keyframes) - 1), \(i) {
  t <- c(
    rep(0, frames_per_hold),
    ease(seq(0, 1, length.out = frames_per_move + 1)[-1])
  )
  tibble(
    b0 = keyframes$b0[i] + t * (keyframes$b0[i + 1] - keyframes$b0[i]),
    b1 = keyframes$b1[i] + t * (keyframes$b1[i + 1] - keyframes$b1[i])
  )
}) |>
  list_rbind() |>
  mutate(frame = row_number())

candidate_resids <- cross_join(candidate_lines, ols_sample) |>
  mutate(
    fitted = b0 + b1 * distance_mi,
    resid = pb_ppm - fitted,
    sq_xmin = distance_mi,
    sq_xmax = distance_mi + abs(resid) * ratio,
    sq_ymin = pmin(pb_ppm, fitted),
    sq_ymax = pmax(pb_ppm, fitted)
  )
ssr_by_frame <- candidate_resids |>
  summarize(ssr = sum(resid^2), .by = frame) |>
  pull(ssr)

x_lim <- range(ols_sample$distance_mi) + c(-0.1, 0.6)
y_lim <- c(0, max(ols_sample$pb_ppm) + 50)

ssr_anim <- ggplot(candidate_resids, aes(x = distance_mi, y = pb_ppm)) +
  geom_rect(
    aes(xmin = sq_xmin, xmax = sq_xmax, ymin = sq_ymin, ymax = sq_ymax),
    fill = "firebrick",
    alpha = 0.2
  ) +
  geom_abline(
    aes(intercept = b0, slope = b1),
    candidate_lines,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  geom_segment(
    aes(xend = distance_mi, yend = fitted),
    color = "firebrick",
    linewidth = 1.2
  ) +
  geom_point(shape = 21, size = 3, fill = "white") +
  coord_fixed(ratio = ratio, xlim = x_lim, ylim = y_lim, expand = FALSE) +
  labs(
    title = "SSR = {scales::comma(ssr_by_frame[as.integer(current_frame)])}",
    x = "Distance (mi)",
    y = "Soil Pb (ppm)"
  ) +
  gganimate::transition_manual(frame)
gganimate::animate(
  ssr_anim,
  nframes = max(candidate_lines$frame) + 40,
  fps = 10,
  end_pause = 40,
  renderer = gganimate::gifski_renderer()
)
```

# Interpret the parameters of a regression model

## The model summary

```{r}
#| echo: true
#| class-output: small-output

# all_samples_north = all samples north of Exide
soil_lm <- lm(pb_ppm ~ distance_mi, all_samples_north)

summary(soil_lm)
```

## The coefficient estimates $\beta_0, \beta_1$

$$
\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon \\
\epsilon &\sim Normal(0, \sigma)
\end{align}
$$

```{r}
# Coefficient table formatted like summary.lm(), with the estimates highlighted
coef_tbl <- coef(summary(soil_lm))
p_vals <- coef_tbl[, "Pr(>|t|)"]
tibble(
  ` ` = rownames(coef_tbl),
  Estimate = sprintf("%.2f", coef_tbl[, "Estimate"]),
  `Std. Error` = sprintf("%.2f", coef_tbl[, "Std. Error"]),
  `t value` = sprintf("%.3f", coef_tbl[, "t value"]),
  `Pr(&gt;|t|)` = format.pval(p_vals, digits = 3, eps = .Machine$double.eps)
) |>
  knitr::kable(format = "html", escape = FALSE, align = "lrrrrl") |>
  kableExtra::kable_styling(full_width = FALSE, font_size = 28) |>
  kableExtra::row_spec(
    0,
    monospace = TRUE,
    extra_css = "white-space: nowrap;"
  ) |>
  kableExtra::column_spec(1:5, monospace = TRUE) |>
  kableExtra::column_spec(2, bold = TRUE, background = "#FFF3B0")
```

## The standard errors $SE(\beta)$

$$
\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon \\
\epsilon &\sim Normal(0, \sigma)
\end{align}
$$

```{r}
# Coefficient table formatted like summary.lm(), with the estimates highlighted
coef_tbl <- coef(summary(soil_lm))
p_vals <- coef_tbl[, "Pr(>|t|)"]
tibble(
  ` ` = rownames(coef_tbl),
  Estimate = sprintf("%.2f", coef_tbl[, "Estimate"]),
  `Std. Error` = sprintf("%.2f", coef_tbl[, "Std. Error"]),
  `t value` = sprintf("%.3f", coef_tbl[, "t value"]),
  `Pr(&gt;|t|)` = format.pval(p_vals, digits = 3, eps = .Machine$double.eps)
) |>
  knitr::kable(format = "html", escape = FALSE, align = "lrrrrl") |>
  kableExtra::kable_styling(full_width = FALSE, font_size = 28) |>
  kableExtra::row_spec(
    0,
    monospace = TRUE,
    extra_css = "white-space: nowrap;"
  ) |>
  kableExtra::column_spec(1:5, monospace = TRUE) |>
  kableExtra::column_spec(3, bold = TRUE, background = "#FFF3B0")
```

## The residual standard deviation $\sigma$

$$
\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon \\
\epsilon &\sim Normal(0, \sigma)
\end{align}
$$

```{r}
#| echo: true
sigma(soil_lm)
```

## $\sigma$ visualized

```{r}
#| fig-width: 12
#| fig-height: 5
#| fig-align: center
# Pick 50 soil samples whose residuals approximate a normal distribution:
# match each standard normal quantile to the nearest unused residual
soil_sigma <- sigma(soil_lm)
soil_resid <- all_samples_north |>
  mutate(
    fitted = fitted(soil_lm),
    resid = residuals(soil_lm),
    sigmas = cut_width(resid / soil_sigma, 0.5)
  )
targets <- qnorm(ppoints(50)) * soil_sigma
picked <- integer(0)
for (target in targets) {
  dist_to_target <- abs(soil_resid$resid - target)
  dist_to_target[picked] <- Inf
  picked <- c(picked, which.min(dist_to_target))
}
normal_sample <- soil_resid[picked, ] |>
  filter(distance_mi < 2)

resid_scatter <- ggplot(normal_sample, aes(x = distance_mi, y = pb_ppm)) +
  geom_abline(
    intercept = coef(soil_lm)[1],
    slope = coef(soil_lm)[2],
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  geom_segment(
    aes(xend = distance_mi, yend = fitted),
    color = "firebrick",
    linewidth = 0.8
  ) +
  geom_point(shape = 21, size = 2.5, fill = "white") +
  labs(x = "Distance (mi)", y = "Soil Pb (ppm)")

# Bins are half a sigma wide so bin edges line up with 1 and 2 sigma
max_count <- max(table(floor(normal_sample$resid / (soil_sigma / 2))))
resid_hist <- ggplot(normal_sample, aes(x = resid)) +
  geom_histogram(
    binwidth = soil_sigma / 2,
    boundary = 0,
    fill = "firebrick",
    color = "white",
    alpha = 0.6
  ) +
  geom_vline(xintercept = 0, linewidth = 1) +
  annotate(
    "segment",
    x = 0,
    xend = soil_sigma,
    y = max_count + 1,
    yend = max_count + 1,
    linewidth = 1.2,
    arrow = arrow(ends = "both", length = unit(0.1, "in"))
  ) +
  annotate(
    "text",
    x = soil_sigma / 2,
    y = max_count + 1,
    label = "sigma",
    parse = TRUE,
    vjust = -0.4,
    size = 9
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = "Residual (ppm)", y = "Count")

patchwork::wrap_plots(resid_scatter, resid_hist)
```

# Recap

## Say hello to the linear model

The linear model is the building block of regression modeling. 

Every model you'll see in this course from now on will build on this.

$$
\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon \\
\epsilon &\sim Normal(0, \sigma)
\end{align}
$$

## A linear model describes a relationship

```{r}
#| fig-width: 6
#| fig-align: center
ggplot(all_samples_north, aes(x = distance_mi, y = pb_ppm)) +
  geom_point(shape = 21, alpha = 0.25) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = "cornflowerblue",
    linewidth = 1.5
  ) +
  labs(x = "Distance (mi)", y = "Soil Pb (ppm)")
```

## Components of the linear model

- Coefficients
- Residuals
- Residual standard deviation

$$
\begin{align}
y &= \beta_0 + \beta_1 x + \epsilon \\
\epsilon &\sim Normal(0, \sigma)
\end{align}
$$

## Interpreting the linear model

```{r}
# Coefficient table formatted like summary.lm(), with the estimates highlighted
coef_tbl <- coef(summary(soil_lm))
p_vals <- coef_tbl[, "Pr(>|t|)"]
tibble(
  ` ` = rownames(coef_tbl),
  Estimate = sprintf("%.2f", coef_tbl[, "Estimate"]),
  `Std. Error` = sprintf("%.2f", coef_tbl[, "Std. Error"]),
  `t value` = sprintf("%.3f", coef_tbl[, "t value"]),
  `Pr(&gt;|t|)` = format.pval(p_vals, digits = 3, eps = .Machine$double.eps)
) |>
  knitr::kable(format = "html", escape = FALSE, align = "lrrrrl") |>
  kableExtra::kable_styling(full_width = FALSE, font_size = 28) |>
  kableExtra::row_spec(
    0,
    monospace = TRUE,
    extra_css = "white-space: nowrap;"
  ) |>
  kableExtra::column_spec(1:5, monospace = TRUE)
```

$$
\text{Pb (ppm)} = 325.6 - 46.1 \times \text{Distance (mi)} + \epsilon
$$
