---
title: "Lab 3: Simulate data and fit a regression model"
subtitle: "Ordinary least squares"
format:
  html:
    other-links:
      - text: Download Source
        href: lab3.qmd
execute:
  eval: false
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
set.seed(222)
theme_set(theme_bw())
```

## Before you code

Before you start coding in this document, your group should have completed the sections **Design the scenario** and **Choose your parameters** from the handout.

## Simulate the data

You've chosen your parameters ($\beta_0$, $\beta_1$, and $\sigma$). Define them here:

```{r}
beta_0 <- ???
beta_1 <- ???
sigma <- ???
```

### Generate random values

1. Simulate your predictor variable.

::: {.callout-tip}

You may use any random number generating function to generate your predictor variable. Often, it's simplest to simulate the predictor variable using the _uniform_ distribution. Here's an example of generating 5 random numbers between 1 and 2:

```{r}
runif(n = 5, min = 1, max = 2)
```

:::

```{r}
# Edit the following code to simulate your _predictor_ variable
your_predictor_variable <- some_random_number_generating_function(
  n = your_sample_size,
  any_other_function_parameters_go_here
)
```

2. Calculate the expected value of the response.

```{r}
# Edit the following code to calculate the _expected_ value of the response
expected_response <- what + goes * here #?
```

3. Simulate the response variable.

```{r}
# Edit the following code to simulate your _response_ variable
your_response_variable <- rnorm(
  n = your_sample_size,
  mean = the_expected_value_of_the_response,
  sd = your_sigma_parameter
)
```

Visualize your simulated data. Verify it's comparable to your original drawing. If necessary, tweak your parameter values until it represents your scenario reasonably well.

```{r}
# First, put your predictor and response variables into a single data frame.

# Then, use ggplot to make a scatter plot.
```

## Fit and interpret a regression model

### Fit the model

Use `lm()` to fit a regression model to your simulated data. 


::: {.callout-tip}

`lm()` expects a _formula_ stating your predictor and response variables. It should look like: `response ~ predictor`.

:::


```{r}
your_model <- lm(your_model_formula, data = your_simulated_data)
# Use summary() to get the model coefficient estimates and their standard errors
# Use sigma() to get the model's sigma estimate
```

### Interpret the model

Refer back to the worksheet to complete most of this section. When you reach the end ("Test your intuition by modifying your simulation to increase $\sigma$ and n independently, then re-run it several times."), come back to this document and use the following code chunks to test your intuition.

**Increase $\sigma$**

```{r}
# Keep your predictor variable the same, but increase sigma by 2x
double_sigma <- ???
  # Modify the following lines of code to re-run your simulation with the bigger sigma
  expected_response <- what + goes * here
#?
your_response_variable <- rnorm(
  n = your_sample_size,
  mean = the_expected_value_of_the_response,
  sd = your_sigma_parameter
)
your_model <- lm(your_model_formula, data = your_simulated_data)
# Get your betas and their standard errors

# How do they compare to the population parameters, relative to the original simulation?
```

**Increase n**

```{r}
# Now increase n by 2x. You'll need to create more predictor variable values.
double_n <- ???your_predictor_variable_2n <- c(
  your_predictor_variable,
  additional_predictor_variable_values
)

# Modify the following lines of code to re-run your simulation with the bigger n
expected_response <- what + goes * here #?
your_response_variable <- rnorm(
  n = the_bigger_sample_size,
  mean = the_expected_value_of_the_response,
  sd = your_sigma_parameter
)
your_model <- lm(your_model_formula, data = your_simulated_data)
# Get your betas and their standard errors

# How do they compare to the population parameters, relative to the original simulation?
```
